Risk Factors and Treatment of Refractory Anaphylaxis - a review of case reports

### 2018-02-09 analysis/paper/paper.Rmd
changed the template

---
title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review'
author:
  - Wojciech Francuzik
  - Sabine Dölle
  - Margitta Worm, MD
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output: 
  bookdown::word_document2:
    fig_caption: yes
    reference_docx: "../templates/template.docx"
bibliography: references.bib
csl: erci.csl
abstract: |
  **Introduction:** Patients experiencing anaphylaxis who do not recover after the treatment with intramuscular adrenaline are regarded to suffer from refractory anaphylaxis. The incidence of refractory anaphylaxis is estimated to range between 3--5% of anaphylaxis cases. The risk factors for refractory anaphylaxis are unknown.

@@

# Introduction
Anaphylaxis is a potentially life-threatening, primarily mast cell-dependent reaction to most frequently food, drugs, insect venom. Anaphylaxis is a clinical diagnosis. It is highly likely if there is an acute onset of mainly skin symptoms in conjunction with cardiovascular symptoms like hypotension and/or respiratory compromise [@Sampson2006]. The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].

Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], a standardized set of criteria to diagnose anaphylaxis have been established [@Sampson2006]. By contrast, the term refractory anaphylaxis has not been established yet. For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of a minimum 300 µg adrenaline. Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e. rapid increase in blood pressure, bronchodilatation, tachycardia).

@@

We performed a database search in the "PubMed" database using the "RISmed"[@rismed] package for R statistical software [@R]. The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017. The search was conducted on 16.11.2017 and returned 150 results. All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenaline were included in this study. Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded. Additionally, case reports referenced in the included sources were added. We excluded reviews and articles not containing a case description as well as cases which documented an appropriate response to injected adrenaline. Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included. We finally included `r unique(d$Sources) %>% length()` Papers reporting `r d$Author %>% length()` cases of anaphylaxis. The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).

We collected information about the elicitors, clinical setting, concomitant diseases, medication, reaction symptoms, therapy and outcomes of the treatment. We used piloted polls to gather the data and all reports were screened once by the same researcher. We investigated which interventions were reported as successful in the treatment of anaphylactic hypotension that was not responding to injected (i.v. or i.m.) adrenaline. Successful and unsuccessful management of anaphylaxis was one of the most important outcomes we evaluated. This information was available in all cases. All of the studies were case reports, therefore, the evidence level was classified as low. The review protocol was not established prior to the review and the review did not undergo preregistration.

# Risk factors for refractory anaphylaxis

@@

# Therapy of refractory anaphylaxis

@@

# Methylene blue for the treatment of refractory hypotension

@@

The hypotension in anaphylaxis is mediated mainly through histamine [@Enjeti1983] which stimulates the endothelial nitric oxide synthase (eNOS) to produce nitric oxide (NO). NO subsequently activates the soluble guanylate cyclase (sGC) which is responsible for the production of cGMP out of guanosine triphosphate (GTP) [@Lucas2000]. This leads to the an opening of the calcium ion channels in smooth muscle tissues and the subsequent loss of vascular muscle contractility [@Omar2015].

@@

# Vasopressors

@@

The use of vasopressin in refractory anaphylaxis was only suggested in the management recommendations from the AAAAI [@Lieberman] and was never vastly investigated. Dopamine, however, is recommended by 2 out of three guidelines. Nevertheless, dopamine might be related to higher a risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015]. Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension. Vasopressin acts on the V1 receptors, while norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].

@@

# Glucagon in anaphylaxis with concomitant beta blockers

@@

# Stopping the elicitor exposure
In cases where the elicitor of anaphylaxis is known, and first-line therapy fails, it is advisable to stop the exposition to the likely allergen. Three case reports described the rapid improvement of a refractory anaphylaxis after preventing further contact with the allergen.

@@

# Extracorporeal life support

@@

# Conclusion

@@

In clinical practice, it is necessary to identify patients at risk. Such risk factors should particularly be assessed in patients undergoing medical diagnostic and therapeutic measures. The diagnosis of anaphylaxis is based on unspecific clinical presentation, most frequently with hypotension, bronchospasm and skin flushing. In cases where anaphylaxis is possible but the symptoms are not clearly apparent introduction of a point-of-case test could confirm the suspition in order to introduce immediate therapy with adrenaline. The measurement of tryptase enables to identify patients who are at risk for severe anaphylaxis (specificity of 0.93) but its sensitivity of 0.36 is insufficient [@Brown2004].

@@

```{r wholeTab}
knitr::kable(whole.tab,
             caption = "List of papers and anaphylaxis cases included in the sudy, summarizing their elicitors, and effective therapy interventions")
```

>>>>

---
title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a review of case reports'
author:
  - Wojciech Francuzik
  - Sabine Dölle
  - Margitta Worm, MD
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output: 
  bookdown::word_document2:
    fig_caption: yes
    reference_docx: "../templates/templateNumberedSections.docx"
bibliography: references.bib
csl: erci.csl
abstract: |
  **Introduction:** Patients experiencing anaphylaxis who do not recover after the treatment with intramuscular adrenaline are regarded to suffer from refractory anaphylaxis. The incidence of refractory anaphylaxis is estimated to range between 3--5% of anaphylaxis cases. The risk factors for refractory anaphylaxis are unknown.

@@

**Keywords:** anaphylaxis; epinephrine; glucagon; hypotension; methylene blue; shock

# Introduction
Anaphylaxis is a potentially life-threatening, primarily mast cell-dependent reaction to most frequently food, drugs, insect venom. Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], it is highly likely if any of the following 3 criterita are fulfilled: 1) Acute onset of an illness with generalized skin / mucosal involvemt with flushing, hives or pruritus; 2) acute cutaneous, cardiorespiratory or gastrointestinal symptoms after exposure to a likely allegen; c) acute reduction of blood pressure (BP) after exposure to a known allergen [@Sampson2006]. The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].

By contrast, the term refractory anaphylaxis has not been established yet. For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of a minimum 300 µg adrenaline. Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e. rapid increase in blood pressure, bronchodilatation, tachycardia).

@@

We performed a database search in the "PubMed" database using the "RISmed"[@rismed] package for R statistical software [@R]. The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017. The search was conducted on 16.11.2017 and returned 150 results. All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenaline were included in this study. Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded. Additionally, case reports referenced in the included sources were added. We excluded reviews and articles not containing a case description as well as cases which documented an appropriate response to injected adrenaline. Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included. We finally included `r unique(d$Source) %>% length()` Papers reporting `r d$Author %>% length()` cases of anaphylaxis. The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).

We collected information about the elicitors, clinical setting, concomitant diseases, medication, reaction symptoms, therapy and outcomes of the treatment. We used piloted polls to gather the data and all reports were screened once by the same researcher. We investigated which interventions were reported as successful in the treatment of anaphylactic hypotension that was not responding to injected (i.v. or i.m.) adrenaline. Successful and unsuccessful management of anaphylaxis was one of the most important outcomes we evaluated. This information was available in all cases. All of the studies were case reports, therefore, the evidence level was classified as low. The review protocol was not established prior to the review and the review did not undergo preregistration. It is important to mention that as a review of case reports this study did not use control groups to compare outcomes and its statistical validity is therefore limited.

## Risk factors for refractory anaphylaxis

@@

## Therapy of refractory anaphylaxis

@@

## Methylene blue for the treatment of refractory hypotension

@@

The hypotension in anaphylaxis is mediated mainly through histamine [@Enjeti1983] which stimulates the endothelial nitric oxide synthase (eNOS) to produce nitric oxide (NO). NO subsequently activates the soluble guanylate cyclase (sGC) which is responsible for the production of cGMP out of guanosine triphosphate (GTP) [@Lucas2000]. This leads to the opening of the calcium ion channels in smooth muscle tissues and the subsequent loss of vascular muscle contractility [@Omar2015].

@@

## Vasopressors

@@

The use of vasopressin in refractory anaphylaxis was only suggested in the management recommendations from the AAAAI [@Lieberman] and was never vastly investigated. Dopamine, however, is recommended by 2 out of three guidelines. Nevertheless, dopamine might be related to a higher risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015]. Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension. Vasopressin acts on the V1 receptors, while norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].

@@

## Glucagon in anaphylaxis with concomitant beta blockers

@@

## Stopping the elicitor exposure
In cases where the elicitor of anaphylaxis is known, and first-line therapy fails, it is advisable to stop the exposition to the likely allergen. Three case reports described the rapid improvement of a refractory anaphylaxis after preventing further contact with the allergen.

@@

## Extracorporeal life support

@@

## Conclusion

@@

In clinical practice, it is necessary to identify patients at risk. Such risk factors should particularly be assessed in patients undergoing medical diagnostic and therapeutic measures. The diagnosis of anaphylaxis is based on non-specific clinical presentation, most frequently with hypotension, bronchospasm and skin flushing. In cases where anaphylaxis is possible but the symptoms are not clearly apparent introduction of a point-of-case test could confirm the suspition in order to introduce immediate therapy with adrenaline. The measurement of tryptase enables to identify patients who are at risk for severe anaphylaxis (specificity of 0.93) but its sensitivity of 0.36 is insufficient [@Brown2004].

# Five year View

@@

```{r wholeTab}
knitr::kable(whole.tab,
             caption = "List of papers and anaphylaxis cases included in the study, summarizing their elicitors, and effective therapy interventions")
```
